---
title: "Project 2"
format: 
  html:
    embed-resources: true
    code-tools: true
  pdf: default
editor: visual
---

# Introduction

-   **`animal_type`**: The type of animal admitted to the shelter.

-   **`month`**: Month the animal was admitted, recorded numerically with January=1.

-   **`year`**: Year the animal was admitted to the shelter.

-   **`intake_type`**: Reason for the animal being admitted to the shelter.

-   **`outcome_type`**: Final outcome for the admitted animal.

-   **`chip_status`**: Indicates if the animal had a microchip with owner information.

-   **`time_at_shelter`**: Days spent at the shelter between being admitted and the final outcome.

In this Generalised Linear Model analysis, we will investigate:

-   Which factors influence the number of days an animal spends in the shelter before their final outcome is decided?

# Data Processing

```{r}
# Load the data and display the structure
shelter <- read.csv("dataset19.csv")
str(shelter)
```

```{r}
# Check for missing values
missing_values <- colSums(is.na(shelter))
missing_values
```


```{r}
# Convert categorical variables to 'factor'
shelter$animal_type <- as.factor(shelter$animal_type)
shelter$intake_type  <- as.factor(shelter$intake_type )
shelter$outcome_type <- as.factor(shelter$outcome_type)
shelter$chip_status <- as.factor(shelter$chip_status)
```

# Explanatory Data Analysis

```{r, echo = FALSE, fig.cap="Distribution of Time at Shelter.", fig.align="center"}
# Visualise the distribution of `time_at_shelter`
library(ggplot2)
library(gridExtra)

# Histogram showing distribution
p1 <- ggplot(shelter, aes(x = time_at_shelter)) + 
  geom_histogram(binwidth = 1, color = "white") +
  labs(x = "Days", y = "Frequency") 

# Boxplot showing the distribution's summary statistics
p2 <- ggplot(shelter, aes(x = time_at_shelter, y = "")) + 
  geom_boxplot() +
  labs(x = "Days", y = "")

# Display plots side by side
grid.arrange(p1, p2, nrow = 1, top = "Figure 1. Time at Shelter Analysis")
```


```{r, echo = FALSE, fig.cap="Time at Shelter by Different Categories.", fig.align="center"}
# Compare 'time_at_shelter' distributions across categories
theme <- theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6), plot.title = element_text(size = 10))
# Create individual plots
g1 <- ggplot(shelter, aes(x = animal_type, y = time_at_shelter, fill = animal_type)) + 
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Animal Type") +
  theme

g2 <- ggplot(shelter, aes(x = intake_type, y = time_at_shelter, fill = intake_type)) + 
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Intake Type") +
  theme

g3 <- ggplot(shelter, aes(x = outcome_type, y = time_at_shelter, fill = outcome_type)) + 
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Outcome Type") +
  theme

g4 <- ggplot(shelter, aes(x = chip_status, y = time_at_shelter, fill = chip_status)) + 
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Chip Status") +
  theme

# Arrange plots in a 2x2 matrix layout
grid.arrange(g1, g2, g3, g4, nrow = 1)
```
