---
title: "Data Analysis Skills – Group Project 2"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
  pdf: default
editor_options: 
  chunk_output_type: console
Author: Group 19
---

# Introduction

The dataset contain a variety of information relatingto each animal admitted to the shelter, with following variables, recorded by animal admission:

-   **`animal_type`**: The type of animal admitted to the shelter.

-   **`month`**: Month the animal was admitted, recorded numerically with January=1.

-   **`year`**: Year the animal was admitted to the shelter.

-   **`intake_type`**: Reason for the animal being admitted to the shelter.

-   **`outcome_type`**: Final outcome for the admitted animal.

-   **`chip_status`**: Indicates if the animal had a microchip with owner information.

-   **`time_at_shelter`**: Days spent at the shelter between being admitted and the final outcome.

In this Generalised Linear Model analysis, we will investigate:

-   Which factors influence the number of days an animal spends in the shelter before their final outcome is decided?

# Data Processing

```{r}
# Load the data and display the structure
shelter <- read.csv("dataset19.csv")
str(shelter)
```

```{r}
# Check for missing values
missing_values <- colSums(is.na(shelter))
missing_values
```

After checking, there are no missing values in the dataset. Then converted the categorical variables (**`animal_type`**, **`intake_type`**, **`outcome_type`**, **`chip_status`**) to the appropriate data type.

```{r}
# Convert categorical variables to 'factor'
shelter$animal_type <- as.factor(shelter$animal_type)
shelter$intake_type  <- as.factor(shelter$intake_type )
shelter$outcome_type <- as.factor(shelter$outcome_type)
shelter$chip_status <- as.factor(shelter$chip_status)
```

# Explanatory Data Analysis

First, conduct an EDA to visualise the distribution of `time_at_shelter`.

```{r, echo = FALSE, fig.cap="Distribution of Time at Shelter.", fig.align="center"}
# Visualise the distribution of `time_at_shelter`
library(ggplot2)
library(gridExtra)

# Histogram showing distribution
p1 <- ggplot(shelter, aes(x = time_at_shelter)) + 
  geom_histogram(binwidth = 1, color = "white") +
  labs(x = "Days", y = "Frequency") 

# Boxplot showing the distribution's summary statistics
p2 <- ggplot(shelter, aes(x = time_at_shelter, y = "")) + 
  geom_boxplot() +
  labs(x = "Days", y = "")

# Display plots side by side
grid.arrange(p1, p2, nrow = 1)
```

The histogram and boxplot showed a right-skewed with the presence of several outliers, indicating that while most animals stayed in the shelter for relatively short periods of time, there were a few that stayed for longer periods of time.

Next, explore the relationship between **`time_at_shelter`** and the categorical variables (**`animal_type`**, **`intake_type`**, **`outcome_type`**, **`chip_status`**) .

```{r, echo = FALSE, fig.cap="Time at Shelter by Different Categories.", fig.align="center"}
# Compare 'time_at_shelter' distributions across categories
theme <- theme(axis.text.x = element_text(angle = 25, hjust = 1, size = 6), plot.title = element_text(size = 10))
# Create individual plots
g1 <- ggplot(shelter, aes(x = animal_type, y = time_at_shelter, fill = animal_type)) + 
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Animal Type") +
  theme

g2 <- ggplot(shelter, aes(x = intake_type, y = time_at_shelter, fill = intake_type)) + 
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Intake Type") +
  theme

g3 <- ggplot(shelter, aes(x = outcome_type, y = time_at_shelter, fill = outcome_type)) + 
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Outcome Type") +
  theme

g4 <- ggplot(shelter, aes(x = chip_status, y = time_at_shelter, fill = chip_status)) + 
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Chip Status") +
  theme

# Arrange plots in a 2x2 matrix layout
grid.arrange(g1, g2, g3, g4, nrow = 1)
```

The box plots reveal several insights into how the duration of stay at the shelter (**`time_at_shelter`**) varies across different categories:

-   **Animal Type:** Some variation appears to exist in the median length of stay in the shelter for different animal types.

-   **Intake Type:** Reason for sheltering has strong influences on the time animals spend at the shelter, with noticeable differences between categories.

-   **Outcome Type:** The final outcome for the animal contributes significantly to the length of time an animal stays in the shelter.

-   **Chip Status:** The presence of a microchip may affect the time spent at the shelter, though the effect seems to be insignificant compared to other factors..

Based on these observations, **`animal_type`**, **`intake_type`**, **`outcome_type`**, and **`chip_status`** could be significant predictors of the time an animal spends at the shelter. In our GLM, we will consider these variables as explanatory variables.

# Generalised Linear Model

Given the nature of our response variable (**`time_at_shelter`**), which counts the number of days, a Poisson regression could initially be considered. However, onsidering the clear over-dispersion (indicated by the presence of outliers), a negative binomial regression may be more appropriate. Negative binomial regression will continue to be used to fit the GLM given the skewness and potential over-dispersion of the count data.

```{r}
# Generalised Linear Model (GLM) with Negative Binomial Distribution
library(MASS)
model_nb <- glm.nb(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status, data = shelter)
summary(model_nb)
```

```{r}
model_poisson <- glm(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status, 
                     data = shelter, family = "poisson")
summary(model_poisson)
```

# Model Evaluation
