---
title: "Data Analysis Skills – Group Project 2"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
  pdf: default
editor_options: 
  chunk_output_type: console
Author: Group 19
---



```{r}
# Load the data and display the structure
shelter <- read.csv("dataset19.csv")
str(shelter)
# Check for missing values
missing_values <- colSums(is.na(shelter))
missing_values
# Convert categorical variables to 'factor'
shelter$animal_type <- as.factor(shelter$animal_type)
shelter$intake_type  <- as.factor(shelter$intake_type )
shelter$outcome_type <- as.factor(shelter$outcome_type)
shelter$chip_status <- as.factor(shelter$chip_status)
```

Given the nature of our response variable (**`time_at_shelter`**), which counts the number of days, a Poisson regression could initially be considered. However, considering the clear over-dispersion (indicated by the presence of outliers), a negative binomial regression may be more appropriate. Negative binomial regression will continue to be used to fit the GLM given the skewness and potential over-dispersion of the count data.

```{r}
# Generalised Linear Model (GLM) with Negative Binomial Distribution
library(MASS)
model_nb <- glm.nb(time_at_shelter ~ animal_type + intake_type + outcome_type + chip_status, data = shelter)
summary(model_nb)
```

